---
title: "assignment 5.1"
author: "Drew Murray"
date: "2022-12-18"
output: html_document
---

```{r, echo=TRUE}
wine = read.csv("C:/Users/dgmur/Downloads/wineFl2022.csv")

#1

#a

par(mar = c(1, 1, 1, 1))

par(mfrow=c(3,3))

boxplot(wine$fixed_acidity, horizontal = TRUE)

boxplot(wine$volatile_acidity, horizontal = TRUE)

boxplot(wine$citric_acid, horizontal = TRUE)

boxplot(wine$residual_sugar, horizontal = TRUE)

boxplot(wine$chlorides, horizontal = TRUE)

boxplot(wine$free_sulfur_dioxide, horizontal = TRUE)

boxplot(wine$total_sulfur_dioxide, horizontal = TRUE)

boxplot(wine$density, horizontal = TRUE)

boxplot(wine$pH, horizontal = TRUE)

boxplot(wine$sulphates, horizontal = TRUE)

boxplot(wine$alcohol, horizontal = TRUE)


qq1 = qqnorm(wine$fixed_acidity)
qqline(wine$fixed_acidity)


qq2 = qqnorm(wine$volatile_acidity)
qqline(wine$volatile_acidity)


qq3 = qqnorm(wine$citric_acid)
qqline(wine$citric_acid)


qq4 = qqnorm(wine$residual_sugar)
qqline(wine$residual_sugar)


qq5 = qqnorm(wine$chlorides)
qqline(wine$chlorides)

qq6 = qqnorm(wine$free_sulfur_dioxide)
qqline(wine$free_sulfur_dioxide)

qq7 = qqnorm(wine$total_sulfur_dioxide)
qqline(wine$total_sulfur_dioxide)

qq8 = qqnorm(wine$density)
qqline(wine$density)

qq9 = qqnorm(wine$pH)
qqline(wine$pH)

qq10 = qqnorm(wine$sulphates)
qqline(wine$sulphates)

qq11 = qqnorm(wine$alcohol)
qqline(wine$alcohol)

#all variables besides density show rightly skewed, and according to the q-q plots,
#they lack normality. For density, it shows evidence of being symmetric and normality


apply(wine[1:11],2,shapiro.test)

#using the critical value of 0.995 all variables were signficant in rejecting  
#normality assumption besides pH

#Chi-square plot
wine=as.matrix(wine)
n=nrow(wine)
Xbar=colMeans(wine[,1:11])

S=var(wine[,1:11])

invS=solve(S)

D=rep(0,n)

for (i in 1:n){D[i]=t(wine[i,1:11]-Xbar)%*%invS%*%(wine[i,1:11]-Xbar)}

par(mfrow=c(1,1))
plot(qchisq((seq(1:n)-0.5)/n,11),sort(D),pch=19,main="Chi-square plot",xlab="Chi-square quantiles",ylab="generalized distances",cex.axis=1, cex.lab=1.5, cex.main=1.5)
abline(0,1,lty=1,cex=1.5,col="red")

#B
```
The chi-square plot is used to check normality. Shows deviation from multivariate normality due to outliers skewing the the line upwards. 
```{r}
#C 
 source("C:/Users/dgmur/Downloads/functions.R")
a = distances(wine)
tail(sort(a))

tail(order(a))

round(wine[c(157,63),],2)
```
Observation 157 and 63 are the outliers. 157 is a white wine that has an exteme value in total sulfur dioxide, and 63 is a red wine that has an extreme value free sulfur dioxide and chlorides. 
```{r}
#D

library(MASS)
source("C:/Users/dgmur/Downloads/functions.R")


wine2 = wine[,1:11]

wine2[,3]=wine[,3]+.01

boxcoxestimate(wine2)

boxcoxplot(wine2[,1])

boxcoxplot(wine2[,2])

boxcoxplot(wine2[,3])

boxcoxplot(wine2[,4])

boxcoxplot(wine2[,5])

boxcoxplot(wine2[,6])

boxcoxplot(wine2[,7])

boxcoxplot(wine2[,8])

boxcoxplot(wine2[,9])

boxcoxplot(wine2[,10])

boxcoxplot(wine2[,11])
par(mfrow=c(4,4))
```
Using the boxcox transformation method. It show that most of the variables have improved after the transformation.
#E

if the population mean for both red wines and white wines are different than combining may cause bimodality Bimodality shows a lack of normality in the data
```
```{r}
#Bonus


e = as.matrix(wine[,c(7,11)])
e[1:3,]

barX <- c(mean(e[,1]), mean(e[,2]))
barX
 
S <- var(e)
 
source("ellipseFunctions.R")

par(mfrow=c(1,1), mar=c(4,4,1,1))

plot(e,xlab="Total Sulfur Dioxide", ylab="Alcohol", pch=15, col=2, cex.axis=1.5,cex.lab=1.5)

points(barX[2]~barX[1], col=1, pch=16,cex=1.5)

points(e[1,2]~e[1,1],pch=16,col=5,cex=2)

ellipsem(barX,solve(S),1.39, col=4)

i=1

xi <-e[i,]

sc <- (xi - barX) %*% solve(S) %*% (xi - barX)

sc
 
Inside=0
 
for( i in 1:nrow(e))
    {
      xi <- e[i,]
      sc <- (xi - barX) %*% solve(S) %*% (xi - barX)
      if(sc < 1.39) Inside=Inside+1
      }

Inside

Inside/nrow(e)
```

```{r}
#2 

men = read.table("C:/Users/dgmur/Downloads/T1-9.dat")
women = read.table("C:/Users/dgmur/Downloads/T8-6.dat")

#a

t1 <- men[,c(2:6)]
t2 <- women[,c(2:6)]

d <- t1 - t2

dbar<-colMeans(d)

round(dbar,2)

covd = cov(d)



library(ICSNP)

HotellingsT2(t1,t2, level = 0.5)

#male mean vector is different from female mean vector
#b
p<-ncol(t1)

alpha = 0.05

m = men[,c(1:6)]

w = women[,c(1:6)]

n1 = nrow(m)

n2 = nrow(w)


for(i in 1:p){
  
   cat("Variable",i, dbar[i] + c(-1,1)*qt(alpha/(2*p),n-1,lower.tail=FALSE)
       
       *sqrt(covd[i,i]/n),"\n")
}


L1 = cov(t1)

L2= cov(t2)

Sp=((n1-1)/(n1+n2-2))*L1 + ((n2-1)/(n1+n2-2))*L2

xbar1 = colMeans(t1) 

xbar2 = colMeans(t2)

for(i in 1:p){
   cat("Variable",i,(xbar1-xbar2)[i] + c(-1,1)*qt(alpha/(2*p),n1+n2-2,lower.tail=FALSE)
         *sqrt((1/n1 + 1/n2)*Sp[i,i]),"\n")
}
#C
```
Paired sample approach was used because each population has a naturally pairing with the country

```{r}
#Bonus Points 2
wine1 = as.data.frame(wine)  

#1
m1 <- manova(cbind(fixed_acidity, density) ~ as.factor(quality), data= wine1)

summary.manova(m1,test = c("Wilks"))

#There is a signifcant difference

#2
m2 = manova(cbind(volatile_acidity, alcohol) ~ as.factor(quality), data= wine1)

summary.manova(m2,test = c("Wilks"))

#There is a significant difference, and the answer did not change

#3
```
Shows that there is not a clear distinction, we cannot conclude that we can differentiate between all of the quality scores